A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-22 paper/paper.tex
Add logical relation (inductive ver) to the paper

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
  \begin{equation*}
    \begin{split}
      [-]_= &:: \mathit{Term} \rightarrow \mathit{Nat} \rightarrow \mathcal{P}(\mathit{Term})\\
      [ [[Void]] ]_i &= \emptyset \\
      [ [[Bool]] ]_i &= \{ [[a]] \mid [[a =>+ true]] \text{ or } [[a =>+ false]] \} \\
      [ [[a ~ b : A]] ]_i &= \{ [[p]] \mid [[p =>+ refl]] \text{ and } [[a<=>b]] \}\\
      [ [[Pi A B]] ]_i &= \{ [[b]] \mid \forall [[a]], \text{ if } [[a]] \in [ [[A]] ]_i\text{, then } [[b a]] \in [ [[B {a}]] ]_i \} \\
      [ [[Set j]] ]_i &= \{ [[A]] \mid [[j < i]] \text{ and } [[A]] \in dom([-]_j) \}\\
      [ [[ A ]] ]_i &= \{ [[a]] \mid \exists [[B]], [[a]] \in[ [[B]] ]_i \text{ and } [[A => B]] \}
    \end{split}
  \end{equation*}
  \caption{Specification of the logical relation for \lang}
  \label{fig:logrel}
\end{figure}

In this section, we present the definition of the logical relation as a
partial function.
Figure~\ref{fig:logrel} shows the specification
of the logical relation that we are about to define. The specification
itself does not immediately induce a definition as it is unclear
whether there exists a smallest partial function that satisfies the
specification.

The construction of a function that meets the specification in
Figure~\ref{fig:logrel} can be broken down into two steps. To build an
infinite hierarchy of universes, we first define a recursor that constructs
the interpretation of the universe level $[[n]]$ given a parameterized
interpretation function for universe levels that are strictly less
than $[[n]]$.
Next, we use the well-foundedness of the less than relation to tie the
knot and define the full interpretation function that meets the
specification.
The resulting definition is then shown to be functional and thus
justifies the notation $[ [[A]] ]_i$ since we know there would be an
unambiguous output corresponding to the type $[[A]]$, assuming that
$[[A]]$ is in the domain of the partial function.

\section{Consistency}
\label{sec:logrelproof}

>>>>

\section{Semantic Typing}
\label{sec:logreldep}
\begin{figure}[h]
\begin{equation*}
  \begin{split}
    [[InterpR i A S]] &:= [[ Interp I i A S  ]] \\
    [[I i]] &:= [[exists S , InterpR i A S]]
  \end{split}
\end{equation*}
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}). The
logical relation takes the form $[[Interp I i A S]]$, where $[[I]]$ is a
function that takes a natural number strictly less than $[[i]]$ to a set of
terms, $[[i]]$ is a natural number representing the universe level,
$[[A]]$ is a type, and $[[S]]$ a set of terms.
% Intuitively, the relation $[[Interp I i A S]]$ takes $[[I]]$, an interpretation
% for types from universes smaller than $[[Set i]]$ and builds the types
% that semantically inhabit $[[Set i]]$.
Ignoring $[[I]]$ for a moment, the notation $[[Interp I i A S]]$ reads
that $[[A]]$ is a level-$[[i]]$ type
semantically inhabited by terms from the set $[[S]]$. For example, the
$[[Void]]$ type is not inhabited by any terms as it corresponds to the
empty set, whereas the $[[Bool]]$ type is inhabted by the set of (not
necessarily closed) terms that reduce to either $[[true]]$ or
$[[false]]$. Not all terms have an interpretation since some terms do
not correspond to types and therefore cannot be inhabited. For
example, $[[true]]$ is a boolean value and therefore does not have a
set corresponding to it.

\Rref{T-Pi} requires some special attention. Let's focus first on the
premises of \rref{T-Pi} and ignore the set that corresponds to $[[Pi A
B]]$. For $[[Pi A B]]$ to have a meaning, we
require that there exists some interpretation $[[S]]$ for the input
type $[[A]]$. For each term $[[a]]$ in the set $[[S]]$, we also
require that there exists some set that corresponds to the term $[[B
{a}]]$. The function $[[F]]$ appearing in the premise is simply a
result of skolemization. In fact, we can alternatively formulate the
premise of \rref{T-Pi} as follows:

This alternative definition becomes problematic when we try to fill
out the missing conclusion of the rule. With the existential variant,
we cannot conveniently refer to the witness that corresponds
to the interpretation of the type $[[B { a }]]$ from the premise. It
is tempting to rewrite the conclusion as follows, though this
alternative definition ...

In a fully dependently typed system, we may perform computations at
the type level. Given $[[empty |- a : A]]$, the type $[[A]]$ may not
be in head form but can reduce to the $[[Bool]]$ type. \Rref{I-Red}
allows us to assign a type-level computation a meaning based on what
type value it reduces to eventually.

The addition of \rref{I-Red} requires us to prove several inversion
lemmas about the logical relation. For example, given a proof of

% In the rules for $[[Interp I i A S]]$, we exclude cases where $[[A]]$
% is neutral. Since we only care about consistency, it suffices to focus
% only on terms that evaluate to closed terms.
\section{Consistency}
\label{sec:logrelproof}
